---
title: "project 2"
output: 
  html_document: 
    css: arial_style.css
date: "2024-07-21"
editor_options: 
  markdown: 
    wrap: 72
---

## Question 1:

What % of variation described by 1st principal component

```{r}

# lad montgomery and pickell data set 
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

# variance formula is d^2 / sum (d^2)

# no transformations 
svd1 = svd(edata)
(svd1$d^2 / sum(svd1$d^2))[1] # 0.89

# log2(data + 1)
svd2 = svd(log2(edata + 1))
(svd2$d^2 / sum(svd2$d^2))[1] # 0.97

# process of elimination (answer is 0.89, 0.97, 0.35)
```

## Question 2:

-   perform log2(data + 1) transform

-   subtract rowMeans() from samples

-   set seed to 333

-   use kmeans to cluster samples into two clusters

-   use svd to calculate the singular vectors

-   correlation between first singular vector and sample clustering
    indicator?

```{r question2}

# perform log2(data + 1) transform 
logged_edata = log2(edata + 1)

# subtract rowMeans() 
logged_edata_centered = logged_edata - rowMeans(logged_edata)

# use kmeans to cluster samples into two clusters 
logged_edata_kmeans = kmeans(t(logged_edata_centered), centers = 2)

# use svd to calculate the singular vectors
svd_logged = svd(logged_edata_centered)

# correlation between first singular vector and sample clustering indicator? 
cor.test(
  svd_logged$v[,1], # first singular vector
  logged_edata_kmeans$cluster # clustering indicator 
) # 0.87 
```

## Question 3:

```{r}

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata_bm = exprs(bm)
pdata_bm=pData(bm)

# fit a linear model relating first gene's counts to number of technical replicates
lm_first_gene = lm(edata_bm[1,] ~ pdata_bm$num.tech.reps)

# plot data for gene versus covariate 
plot(edata_bm[1,] ~ pdata_bm$num.tech.reps)
abline(lm_first_gene)

# why doesn't the model fit well? 
# a bunch of samples are clustered at num of technical reps = 2 
```

## Question 4:

```{r}

# fit lm relating first gene's counts to age of person and sex of samples
lm_age_sex = lm(edata_bm[1,] ~ pdata_bm$age + pdata_bm$gender)
plot(edata_bm[1,] ~ pdata_bm$age + pdata_bm$gender)
abline(lm_age_sex)

tidy(lm_age_sex)
 
# value / interpretation of age coeff? 
# age = -23.91. This coefficient means that for each additional year of age, the count goes down by an average of 23.91 for a fixed sex.
```

## Question 5:

```{r}

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

# log2 + 1 transform 
logged_edata = log2(edata+ 1)

# fit regression model to each sample, population is the outcome 
mod = model.matrix( ~ pdata$population)
multiple_regressions = lm.fit(mod, t(edata))

# dimension of residual matrix? effects matrix / coeffs matrix ?
names(multiple_regressions)
dim(multiple_regressions$residuals)
dim(multiple_regressions$effects)
dim(multiple_regressions$coefficients)
```

## Question 6:

```{r}

# examining effects matrix
View(multiple_regressions$effects)
```

## Question 7:

```{r}

library(limma)

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

# fit many regression models to the expression data where age is outcome variable 

# only keep the rows from pdata_bm (the samples/cols we have data for)
pdata_bm = na.omit(pdata_bm) # drop nulls so dimensions match
edata = edata[, rownames(pdata_bm), drop = FALSE]

mod = model.matrix ( ~ pdata_bm$age )
limma_model = lmFit(edata, mod)

# what is the coeff on the 1000th gene 
limma_model$coefficients[1000,]
```

## Question 8:

```{r}

# fit many regression models to the expression data where age is outcome variable 
# outcome variable tissue.type

# looked at options first
pdata_bm$tissue.type # 17 levels 
dim(edata) # 16 levels 
```

## Question 9:

Why is it difficult to distinguish the study effect from the population
effect in the Montgomery Pickrell dataset from ReCount?

```{r}

# just examine one chice at a time

# The effects are difficult to distinguish because each study only measured one population. 
pdata$study
pdata$population
# true => only one population type per study (can see from stagger)
# so you can't seperate the batch effect from population difference effect 
```

## Question 10:

Set the seed using the command set.seed(33353) then estimate a single
surrogate variable using the sva function after log2(data + 1)
transforming the expression data, removing rows with rowMeans less than
1, and treating age as the outcome (hint: you may have to subset the
expression data to the samples without missing values of age to get the
model to fit). What is the correlation between the estimated surrogate
for batch and age? Is the surrogate more highly correlated with race or
gender?

```{r}
# preprocessing the data
set.seed(33353)

# make sure that all the phenodata and edata dimensions match 
pdata_bm_nonulls = na.omit(pdata_bm)
edata_processed = edata[,rownames(pdata_bm_nonulls), drop=FALSE]

# log2 + 1 transform : logged_edata, remove the rows with rowMeans < 1
logged_edata = log2(edata_processed + 1)
new_logged_edata = logged_edata[rowMeans(logged_edata) > 1, ]

# fit model - treating age as the outcome 
mod = model.matrix(~age, data=pdata_bm)
modnull = model.matrix(~1, data=pdata_bm) # only null 
sva_model = sva(edata, mod, modnull, n.sv = 2)

# correlation between batch / age 
cor(sva_model$sv, pdata_bm_nonulls$age) # age 

# corr race
cor(sva_model$sv, as.numeric(pdata_bm_nonulls$race))

# corr gender
cor(sva_model$sv, as.numeric(pdata_bm_nonulls$gender)) # more highly corr by gender
```
