---
title: "R Notebook"
output: html_notebook
---

## Data Types in Bioconductor

-   experimental data: data that we've collected with high throughtput sequencing, etc.

-   metadata: any annotation on the data, tells us information about the data

-   annotation: data from databases that is relevant to the experiment

    -   gene is highly conserved

    -   gene is super expressed in this cell...

    -   etc

-   Try to take raw and unprocessed data –\> preprocessed data -\> results.

-   Data containers:

    -   Raw data is preoprocessed from all differet sources and put into a common data container (like annotation hub for example) which you can analyze to get the results.

## Annotation 

-   How to annotate data from Bioconductor

-   Look at gene affymetrix mircroarray data

    -   you set up dna using probesets

    -   probsets have identifiers with a code

    -   which peice of dna was measured using this probe -\> affymetrix identifier

    -   ENTREZ - ncbi database

    -   the affymetrix identifier is converted into an ENTREZ number.

-   You can look up any gene location on the UCSC browser.

    -   It has a bunch of rows under a genome with experimental data on segments of the genme

    -   experimental data always has certain biases and problems

-   How to annotate stuff in bioconductor

    -   annotation packages: contain annotations from various sources. Very popular for annotating microarrays.

    -   query online resources: UCSC or ENSEMBL

## Expresssion Set

-   expression set: contains data of measuring the expression of genes

    -   expression matrix: these matrics have features (x), samples (y). Each entry is a number that tells you how much each feature was expressed in that particular sample.

    -   metadata matrix / phenotype data : covariates (x) by samples (y)

        -   covariates examples: time of the experiment, data of the experiment, information on the samples...

    -   metadata on the features / feature data : covariates (x) by features (y)

        -   covariates: info on the feature we are measuring: what gene was measured, etc

    -   **The expression matrix, metadata matrix, and feature metdata matrix make the "Expression Set"**

    -   **"eSet" =\> multiple expression matrices**

        -   for example: for a dna methylation experiment, one matrix corresponds to the methylation

```{r}

# actually coding data container 
library(ALL)

data(ALL)
ALL # this is an expression set with 12.6k features, 128 samples 

experimentData(ALL) # Some metadata, two pubmed IDs... 
abstract(ALL) # You can see the abstract 
?ALL # Acute Lymphoblastic Leukemia Data from the Ritz Laboratory => on the side, it opens up a bunch of information

# Biological Goal: Select the number of samples and features you want displayed.

ALL[,1:5] # the first five samples, with all the features
ALL[1:10, 1:5] # 10 samples, 5 features 

# Biological goal: gene expression data 

# what about the gene expression data? 
exprs(ALL)[1:4, 1:4] # the expression matrix. (x: numbers assigned to features, y: sample ids)
# get the sample and feature names for the data 
sampleNames(ALL)
featureNames(ALL)

# Biological goal: Information about the sample

pData(ALL) # dataframe of the (x: samples, y: information)
pData(ALL)$age # access specific parts 
ALL$age # gives you the same thing 
names(pData(ALL)) # all the metadata columns, the (y:inforamtion)
# pdata has a phenodata column 
pData(phenoData(ALL))

phenoData(ALL) # gives you AnnotatedDataFrame object with information about the samples
# you should stick to using pData though 
# differnt from pdata

varLabels(ALL) # another way to list (y: information)

# Biological goal : Informations about the features 

featureData(ALL) # in theory (x: features, y: information), unfortunately, in practice, it's often empty

# Biological goal: Since the feature Data matrix is emtpy, have to get the feature names and get information on the feature names manually. 

ids = featureNames(ALL) # first five feature names 
ids # feature names

# there is a package that will help us 
library(hgu95av2.db) # a bunch of maps that map feature names to actually interpretable information...
as.list(hgu95av2ENTREZID[ids]) # converts the ids into ENTREZ IDs, meaningful identifiers. 
```

## Summarized Experiment 

-   newer version of expression set

```{r}
library(airway)

# laoding the neccessary data in... 
data(airway)
airway # 63677 rows/genes/exons/features, 8 samples.
# similar to expression set, slightly different syntax 

# Biological Goal: Get the data on each sample from the phenotype matrix 

colData(airway) # returns DataFrame, pData, the information on each "row" which is a sample
colnames(airway) # names of each sample
# the feature names 
rownames(airway) # names of each feature, basically each gene/exon, whatever, each row

# Biological Goal: Experiment Data, about the paper or research 

metadata(airway) 

# Biological Goal: get the expression numbers by accessin the "assays"
 
airway$cell # access elements 
airway # check the assays part, and it will have a name 
# get the assay based on the name of the assay
# assay(data, name of the assay to extract)
assay(airway, "counts") # rna - seq count data

# Biological Goal: Access the GRanges list associated with each row/feature 

# getting the exons 
length(rowRanges(airway)) # 63677 features, 1 range for each.
rowRanges(airway) # each row is a gene, each GRange in the granges list gives us the exons in the gene 

# Biological goal: how many total exons 

# elementNROWS -> length for each range
# sum them up (all lengths)
sum(elementNROWS(rowRanges(airway))) # 738009 exons f

# Biological goal: get a list with the start coordinate of each exon.
start(airway) 

# Biological goals: get the genes from airway that overlap some grange 
gr = GRanges("1", range = IRanges(start = 1, end = 10^7)) 
subsetByOverlaps(airway, gr) # the genes inside airways that overlap gr
# gives us info: dimensions are 329 8 meaning there are 329 genes from airway that overlap gr 
# you can get the expression data from assays(1) counts for only these overlapping genes. 
```

## GEOQuery

-   tool for interfacing with GEOQuery

-   GEO is widely used repository, containing gene expression data, epigenetic, etc

-   GEO: dataset(s) from a paper (subdataset for rna sequencing, one for chip sequencing)

    -   each dataset has an accession number. and geo identifier

    -   most people downloading from GEO want all data associated with a publication

```{r}

# load in the GEO Data 
library(GEOquery)

# Biological Goal: Get all the datasets into expression set 

# get datasets into list from accession number 
eList = getGEO("GSE11675")
eList
# download the datasets into an expression set 
eData <- eList[[1]]
class(eData) 

# look at the phenotype data for the sample, what information was collected on the sample
names(pData(eData)) 

# Biological Goal: Download Raw Data 

# get the raw data from the supplementary files
# cell file - binary format that holds raw data for microarray data 
# fastq files - raw data for sequenceing 
eList2 = getGEOSuppFiles("GSE11675")
eList2
class(eList2) # r data frame 
```

## biomaRt 

-   interface between R and biomaRt

-   any database that wants to be online can upload it to biomaRt and then we can grab it

    -   queries the database.

-   ENSEMBL, UniProt, etc.

```{r}

library(biomaRt)

# Step 1: Choose a Database and a Dataset inside the Database 

# Database = Mart
head(listMarts()) # the available marts 

# choose the ensemble database
mart <- useMart("ensembl")
head(listDatasets(mart)) # the datasets in the ensembl database 

ensemble <- useDataset("hsapiens_gene_ensembl", mart) # ensembl keeps updating its version of the genome.

# Biological Goal: Get gene names associated with microarray probe IDs 

values <- c("202763_at","209310_s_at","207500_at") # these are the affymetrix probe IDs 
class(values)

# you run a query using getBM (get biomark)
getBM( 
  attributes = c("ensembl_gene_id", "affy_hg_u133_plus_2"), # whatever cols you want in output
  
  filters = "affy_hg_u133_plus_2", # filter search terms 
    # affy_hg_u133_plus_2 again to seaarch this value
  
  values = values, # the values of identifiers.
  mart = ensemble) # the database you want to query 

# output: you get the corresponding gene id for every probe ID 
attributes


# Biological Goal: List attributes of the dataset 

attributes <- listAttributes(ensemble)
attributes # you can go through this and find the attributes 
head(attributes)
tail(attributes)

# Alpine marmot gene stable ID => you'll get a bunch of genes not from humans, but that's because ensembl gets a human gene ID and convert it into a matching gene ID in a different organism => called homologs

filters <- listFilters(ensemble)
filters # view the filters 


# Biological Goal: Attributes grouped into pages, ask for only attributes on a specific page

attributePages(ensemble)
features_only <- listAttributes(ensemble, page = "feature_page")

# to isolate multiple pages -> query the pages individually and merge them 

# sometimes when you use biomart, you get duplicated rows 
```

## R S4 CLASSES 

-   R s4 classes are used to construct complex data structures.

-   s3 -\> classic object oriented system in R.

-   loading in 2 packages

```{r}
library(GenomicRanges)

# S3 => no type safety.

# you can make any object into any type of class
df <- data.frame(y = rnorm(10), x = rnorm(10))
lm.object <- lm(y ~ x, data = df) # constructing an lm.object 
lm.object 
class(lm)
names(lm.object)

# turn a list into a linear model (lm)
xx = list(a = letters[1:3], b = rnorm(4))
class(xx) = "lm" # turn it into a lm class, even though it makes no sense 

# s4 => type safety. You can define what classes need to have and what can be typecasted to each other 

ALL # the expression dataset. 
# s4 class
class(ALL)
isS4(ALL)

# how to get help for a class 
class?ExpressionSet
?"ExpressionSet-class"
?ExpressionSet # documents the constructor

# coding standards for new class 
# classes have to start with capital letter
# classes need a constructor 

# rare to make an expression set from scratch
new("ExpressionSet") # old type of constructor 

# see the definition of the class 
getClass("ExpressionSet") # each of the "slots' has a class 

# the slots is where the data usually is 

# the way you're not supposed to do it
ALL@annotation # gets you the annotation of the ALL Expression Set 
slot(ALL, "annotation")
# use accesser fucntion - getters basically.
annotation(ALL)
# some of the slots are private, not supposed to be used by functions

# OLD_OBJECT = updateObject(OLD_OBJECT) -> overrides old object, updates to the new class definition

# validObject() checks if it's a valid member of its class -> sometimes takes a long time to run. You might create an invalid object if you change instance variables, slot stuff.
```

## S4 Methods 

```{r}
# s4 methods are simple functions, allow you to apply other functions based on them.

# call different methods based on the type of x 
mimicMethod <- function(x) {
    if (is(x, "matrix"))
        method1(x)
    if (is(x, "data.frame"))
        method2(x)
    if (is(x, "IRanges"))
        method3(x)
}
class(mimicMethod)

as.data.frame
class(as.data.frame) # package 
showMethods("as.data.frame") # the list of possible methods in this package

# get documentation for the methods 
getMethod("as.data.frame", "DataFrame") # get the data frame method
method?"as.data.frame,DataFrame"

# get the documentation for the methods 
??"findOverlaps,Ranges,Ranges-method"
```
