---
title: "annotation hub project"
output: anootation hub project
---

# Homework 1: Annotation Hub Project

## Question 1: Use annotation hub to get data on CpG islands in the human genome. How many islands exists on autosomes (chrom 1-22) 

```{r}

ah = AnnotationHub()
qhs = query(ah, "CpG Islands")
qhs # 57 records. 

islands = qhs[["AH5086"]]# just take the first one 
islands # seqnames holds the information about chromosome number, 28691
class(islands) # GRange
print(seqnames(islands)) # the names of the chromosomes.

# merge the overlapping intervals 
reduced_islands <- reduce(islands) # still 28691

# define the vector of autosomes names. 
autosome <- c(paste("chr", 1:22, sep = ""))
autosome

# only include the seqnames (chromosomes) that are in that autosome vector 
seqlevels(reduced_islands, pruning.mode = "coarse") = autosome
seqlevels(reduced_islands)

# look at the number of observations 
reduced_islands
```

## Question 2: How many CpG islands exist on chromosome 4? 

```{r}

# isolate the stuff from chromosome 4 
reduced_islands
chr4_islands <- reduced_islands[seqnames(reduced_islands) == "chr4"]
chr4_islands # 1031 ranges from islands. 
```

## Question 3: Obtain data from H3K27me3/H1 cell line from the Epigenomics Roadmap Using Annotation Hub. Only count the autosomes (chr 1-22).  How many bases do these reigons cover? 

```{r}

# the h1 cell line -> the h1 line is E003 
qhs2 <- query(ahub, c("H3K4me3", "E003")) # query for the cell line and histone modification
qhs2

# download (we used narrow peaks in lecture, will use those again)
h1_cells = qhs2[["AH29884"]]
h1_cells 

# isolate only the autosomes (use the vector from before)
seqlevels(h1_cells, pruning.mode = "coarse") = autosome
seqlevels(h1_cells) # goes to chr22

sort(h1_cells)

# now take the reduction or union to find the total breadth 
sum(width(reduce(h1_cells))) # take the sum of the widths of all the reigons where any segments exist. 
```

## Question 4: Each region should have a "signalValue" so we will find the mean across all regions on the standard chromosomes (H3K27me3)

```{r}

# the h1 cell line -> the h1 line is E003 
qhs3 <- query(ahub, c("H3K27me3", "E003")) # query for the cell line and histone modification
qhs3

h1_cells_2 <- qhs3[["AH29892"]]
seqlevels(h1_cells_2, pruning.mode = "coarse") = autosome

metadata_h1 <- mcols(h1_cells_2)
metadata_h1
mean(metadata_h1$signalValue)
```

## Question 5: Bivalent regions bound by H3K4me3 and H3K27me3. How many bases on standard chromosomes are bivalently marked? 

```{r}

# find the intersection between both of these regions and then add their lenghts together 

bivalent_reigons <- intersect(h1_cells_2, h1_cells) # only the overlapping reigons between these two 
sum(width(bivalent_reigons))
```

## Question 6: To what extent do bivalent reigons overlap the CpG islands? how big a fraction (expressed as a number between 0 and 1) of the bivalent regions, overlap one or more CpG Islands? AKA: How many (number) of bivalent regions out of the total number of bivalent reigons overlap cpg islands.

```{r}

# find the overlaps between them. 
biv_ov <- findOverlaps(bivalent_reigons, reduced_islands) # 22182 in the query overlap with 26641 in the subject 

query_hits <- length(unique(queryHits(biv_ov))) # total hits 
# percentage = 54 percent 
query_hits / length(bivalent_reigons)
```

## Question 7: What percentage of the bases part of the CpG islands overlap the bivalent marked? AKA How many bases from the cpg islands overlap bivalent reigons. 

```{r}
# find the reigons (bases) of interesection only, must be common to both 
cpg_ov <- intersect(reduced_islands, bivalent_reigons) 
cpg_ov # 12174 overlaps 

intersections_width <- sum(width(reduce(cpg_ov))) # how much space do these intersections take up total.

total_width <- sum(width(reduce(reduced_islands))) # total width of the cpg islands 

intersections_width / total_width # intersections bases /bases of cpg islands
```

## Question 8: How many bases are bivalently marked within 10kb of CpG islands? Use the resize() function

```{r}

# first, extend the range of each interval in the cpg islands frame to stretch 10kb in either direction (pushing out from the center)
cpg_stretched <- resize(
  reduced_islands, 
  width = 20000 + width(reduced_islands), 
  fix = "center"
)

# find out how much of this +/- 10kb range overlaps with bivalent reigons
cpg_bivalent <- intersect(cpg_stretched, bivalent_reigons)

# figure out how many bases this is when looking at just the regions of intersection by summing the width of all regions 
sum(width(reduce(cpg_bivalent)))
```

## Question 9: How big a fraction (expressed as 0 and 1) of human genome is in a CpG Island? 

```{r}

# total length of all the cpg islands / length of human genome 

genome_length <- sum(as.numeric(seqlengths(islands)))

sum(seqlengths(reduced_islands)) / 3*10^9

# my answer isn't an option, so i just guessed, no idea what i did wrong.

```

## Question 10: Odds Ratio For Overlap of Bivalent Marks With CPG islands 

```{r}

# make the impty matrix 
inOut = matrix(0, ncol = 2, nrow = 2) # initializes to 0.
colnames(inOut) = c("in", "out") # 
rownames(inOut) = c("in", "out")
inOut 

# bivalent vs cpg islands 

inOut[1,1] = sum(width(intersect(bivalent_reigons, reduced_islands, ignore.strand = TRUE)))
# in bivalent reigons, not in reduced islands 
inOut[1,2] = sum(width(setdiff(bivalent_reigons, reduced_islands, ignore.strand = TRUE)))
inOut[2,1] = sum(width(setdiff(reduced_islands, bivalent_reigons, ignore.strand = TRUE)))
inOut[2,2] = 3 * 10^9 - sum(inOut)

inOut

odds_ratio = inOut[1,1] * inOut[2,2] / (inOut[2,1] * inOut[[1,2]])
odds_ratio # closest to 169, maybe differing genome size? 
```
